A Zero-Knowledge Protocol for Keystroke Authentication

### 2014-12-16 paper/paper.tex
final draft

\author{Noam Zilberstein and Jonathan Chen\thanks{Advised by Nadia Henninger} \\
University of Pennsylvania
}

\maketitle

@@

\section{Introduction}
Using passwords for authentication is an ancient concept that continues to be the commonly accepted standard. For the sake of convenience, many people choose low-strength passwords and reuse them for multiple websites~\cite{Florencio:2007:LSW:1242572.1242661}. This is extremely insecure and begs the question: is there a better way to authenticate \textit{lazy} users?

Several fixes to this problem are already in place and work by augmenting the password with a secondary authentication factor. Websites that suport two factor authentication include Amazon, Facebook, Google, Dropbox, and many others~\cite{2fac}. In each case, the first factor is a password and the second factor is some kind of trusted hardware such as a personal computer or smartphone. When users try to log in from an untrusted machine, they must confirm their identity by typing a code that is sent to a trusted source such as an SMS, email, or phone call. While this process greatly improves the security of a poorly chosen password, it is still a hastle for the lazy user. It also fails if the user does not have access to a trusted piece of hardware. The solution presented in this paper uses trusted hardware that users have at all times: their hands.

@@

A problem arises with this protocol in the case that the database containing users' keystroke data is compromised. In a password model, if a user's password is leaked, then he can simply create a new one. However, in an authentication scheme based on keystroke dynamics, if an adversary gains knowledge of keystroke data, then the affected users can never confidently use keystroke dynamics to authenticate themselves again. In order to fix this, as little data as possible should be stored in the database. In fact, the protocol we present is zero-knowledge. This means that a user will never at any point in the enrollment or authentication process jeopardize the secrecy of his keystroke timings.

@@

\subsection{Keystroke Dynamics}
\label{sec:bg-keystroke}
Long before computers were invented, written signatures were used to verify people's identities. Signatures are commonly used to confirm the acceptance of contracts and to verify credit card and bank transactions. Typing one's name into a computer is the digital age's moral equivalent to writing it down on a piece of paper. It is not considered a breach of security to see someone's signature witten on a piece of paper; the security assumption is that another person cannot copy it to a sufficient level of precision. Similarly, people types their names with high enough frequency that they develop a muscle memory for this process~\cite{monrose:authentication}.

@@

\begin{quote}\small
Peggy would like to convince Victor that two cards are different colors, however Victor is color blind. First, Victor holds up the two cards and asks Peggy to identify the color of each one. Victor then labels the back of the two cards with Peggy's responses, shuffles them behind his back and then asks Peggy to again identify the color of each one. If the cards indeed have different colors, then Peggy can easily identify them. However, if the cards are the same color then she can only give an answer that is consistent with the labels with probability $\frac12$. By repeating this process several times, Victor can be sure that Peggy is not lying. However, Victor cannot recreate the proof himself because he is color blind.
\end{quote}

\noindent Relating this to an authentication protocol, the user can be viewed as a prover who knows some secret and the server is a verifier that must confirm that the user knows the secret, but has no access to the secret itself. In the case of keystroke dynamics, the secret is the user's keystroke timings. In a sense users do not ``know'' their timings, but as stated in Section~\ref{sec:bg-keystroke}, they can consistently produce them. If users know the secret then they should always be able to prove their identities, but with high probability an imposter should not be able to prove knowledge of the secret. These conditions relate directly to the example above. If Peggy were cheating, then she would only have a small probability of guessing the labels correctly many times. In an authentication scheme, this failure probability should be negligably small. In other words, the probability that anyone can prove that they are a certain user without knowing that user's secret is negligable.

@@

However, if the prover does not know the secret, then it would have to produce an alternate value for $c$ such that $g^c\equiv g^{r+bx}\mod p$. The probability of being able to do this is negligable. The protocol is zero-knowledge because, by the discrete log assumption, an adversary can gain no information about $r$ from the value $g^r$, and without $r$ it is impossible to determine $x$ from the value $r + xb$.

@@

\subsection{Fuzzy Extractors}
\label{sec:bg-fuzzy}
Fuzzy extractors are used to extract a uniformly random string $S$ from a noisy input $\mathfrak{B}$, also known as a biometric template. A similar input $\tilde{\mathfrak{B}}$ will product the exact same string $S$~\cite{conf/csreaSAM/AlvarezEA09}. In the case of keystroke based authentication, $S$ is a user's unique signature which can be extracted from a noisy input. It is then possible to prove knowledge of the secret value $S$ using a zero-knowledge proof. The following construction for a fuzzy extractor is due to \'Alvarez, Encinas, Hern\'andez, and Sanchez-Avila.

@@

\begin{figure}[h!]
\centering
\begin{tikzpicture}[node distance = .5cm and 1cm]
    % Place nodes
    \node [block] (msg) {Message};
    \node [block, right= 3cm of msg] (enc) {Encoded Message};
    \node [decision, right= of enc] (plus) {$-$};
    \node [below= of enc] (hash) {$H$};
    \node [below= of plus] (delta) {$\Delta$};
    \node [above= of plus] (temp) {$\mathfrak{B}$};
    \node [block, above left= of msg] (poly) {$p(x)$};
    \node [left= of poly] (secret) {Secret ($S$)};
    \node [below left= of msg] (rand1) {};
    \node [left= of rand1] (rand) {Random Inputs};
    % Draw edges
    \path [line] (secret) -- (poly);
    \path [line] (rand) -| (poly);
    \path [line] (poly) -| node [midway, above] {$y_1y_2\ldots y_n$} (msg);
    \path [line] (rand) -| node [midway, below] {$x_1x_2\ldots x_n$} (msg);
    \path [line] (msg) -- node [midway, above] {\tiny Reed-Solomon Code} (enc);
    \path [line] (enc) -- (hash);
    \path [line] (enc) -- (plus);
    \path [line] (plus) -- (delta);
    \path [line] (temp) -- (plus);
\end{tikzpicture}
\caption{Enrollment phase for a Fuzzy Extractor}
\label{fig:enroll}
\end{figure}

@@

\begin{figure}[h!]
\centering
\begin{tikzpicture}[node distance = .5cm and .75cm]
    % Place nodes
    \node (temp) {$\tilde{\mathfrak B}$};
    \node [decision, right= of temp] (plus) {$+$};
    \node [above= of plus] (delt) {$\Delta$};
    \node [below = of plus] (reject) {Reject};
    \node [decision, right = of plus] (decide) {$\stackrel?=$};
    \node [above = of decide] (h) {$H$};
    \node [block, right= 2cm of decide] (dec) {Decode $(x_i,y_i)$};
    \node [block, right= 1.5cm of dec] (coeff) {$p(x)$};
    \node [right= of coeff] (s) {$S$};
    % Draw edges
    \path [line] (temp) -- (plus);
    \path [line] (delt) -- (plus);
    \path [line] (plus) -- node [midway, above] {$\mathfrak h$} (decide);
    \path [line] (h) -- (decide);
    \path [line] (decide) -- node [midway, below] {\tiny Reed-Solomon}
      node [near start, above] {\tiny yes} (dec);
    \path [line] (decide) |- node [near start, right] {\tiny no} (reject);
    \path [line] (dec) -- node [midway, above] {\tiny Interpolate} (coeff);
    \path [line] (coeff) -- (s);
\end{tikzpicture}
\caption{Extraction phase for a Fuzzy Extractor}
\label{fig:extract}
\end{figure}

\'Alvarez, Encinas, Hern\'andez, and Sanchez-Avila showed that the above method of fuzzy extraction has a genuine acceptance rate of 89\% and a false acceptance rate of around 1\% on iris scan authentication~\cite{conf/csreaSAM/AlvarezEA09}. While a different biometric such as keystroke dynamics would undoubtably produce different error rates, prior work shows that this fuzzy extractor has potential to accurately distinguish between people. The genuine acceptant rate of 89\% is fairly good; it implies that the expected number of trials people will need to perform in order to authenticate themselves is $\frac{100}{89}\approx 1.12$. In the context of the keystroke dynamics protocol, this is convenient enough to satisfy a lazy user. In most instances, the user will only have to type his name once, but occasionally a second trial may be required.

@@

We wish to examine the feasibility of keystroke based authentication from both a theoretical and empirical perspective. Two protocols are presented. The first procotol is extremely insecure, but acts as a proof of concept and comparison baseline in terms of accuracy and security. This protocol has been implemented as a simple web app. The second protocol is a zero-knowledge version of the first one that relies on an extractor to obtain a secret from a noisy input. The secret is never stored or communicated across a network; instead knowledge of the secret is proven using a zero-knowledge protocol. This protocol has been implemented using a rounding extractor and the feasibility of using the fuzzy extractor that is described in Section~\ref{sec:bg-fuzzy} is also examined.

When discussing these protocols, the words \textit{prover} and \textit{client} are used almost interchangeably as are the words \textit{verifier} and \textit{server}. The client is an implementation of a prover and the server is an implementation of a verifier. The client is naturally more `powerful' than the server as the client knows the secret keystroke timings (or can at least extract them from a noisy input). Both the regular and the zero-knowledge versions of the web app have two capabilites: account creation and login. The account creation phase requires users to type their names in a text box. The keystroke timings are recorded and communicated to the server. The server then stores some data about the keystrokes in a database. When users try to log in, they also simply types their names in a text box. The names that are typed are looked up in the database and then their keystroke timings are compared against the ones that are stored. Note that the word \emph{compare} is used loosely here; in the zero-knowledge protocol, no direct comparison occurs per se. Instead, the difference between the two signatures is measured by some specialized technique. The user experience is identical (aside from login accuracy) in the regular and zero-knowledge versions, although the implementation is quite different.

@@

\subsection{The Zero-Knowledge Protocol}
\label{sec:zkp-prot}
The zero-knowledge protocol builds on top of the basic protocol with the added constraint that no meaningful information about the keystroke vector is ever transferred accross the network. This also means that the representation of the keystroke vector that is stored in the database does not leak any information about the keystroke vector itself. If the database were completely compromised, there would thus be no breach in security; the information in the database can be used to verify a user's identity and nothing more.

@@

The authentication protocol is shown in Figure~\ref{fig:zk_auth}. Since the client in this implementation is a web page that may be accessed from any computer, information about a particular user cannot be stored and thus the value of $g$ must be provided by the server after discovering the user's name. Aside from this caveat, the procotol is identical to the zero-knowledge proof presented in Section~\ref{sec:bg-zkp}.

@@

Table \ref{table:res-basic} contains the results from this experiment en the Basic Protocol implementation that is described in Section~\ref{sec:basic-prot}. Trials were conducted between 20 and 40 times with varying word lengths. The values of the timings were compared directly, allowing a total error of $\epsilon=50$. The value $\epsilon'$ in Table~\ref{table:res-basic} indicates the average error value that the target user scored. Some trends arise from these results. As word length increases, typing signatures become more unique and therefore adversary forgery success rate drops significantly. However, this also affects target users' ability to recreate their typing signature consistently. There appears to be a tradeoff between security and convenience which is due to choosing a constant $\epsilon$. Longer keystroke vectors amortise the allowed error over a larger number of keystrokes, therefore each keystroke must be more exact.

@@

Table~\ref{table:res-zkp} documents repeated test trials with $\delta=0.05$. Trials were conducted between 20 and 40 times with varying word lengths. The trend of increasing successful authentications with increasing word length is apparent. This is most likely attributed to the fact that typing patterns are more distinct with longer pieces of text. Conversely, imposters are also more likely to successfully impersonate the target user with a longer text, since the scheme does not measure the accumulated difference in time. In other words, an imposter's speed could be very far from the expected timings on many of the strokes, but close on others. In this case, the imposter may still be authenticated because the protocol only cares \textit{how many} matches occured, not \textit{how much} the totals differ. With shorter words, the implementation is very unforgiving.  Since there are fewer keystrokes to average out the mistakes, timing is extremely sensitive.  In practice, this is less of a concern since virtually no one has a combined first name and last name of less than three characters. There exists a non-trivial difference between the success rates of target users and their impersonators. This affirms the fact that keystroke dynamics are variable enough and have sufficient entropy to be used as authentication tools. Barring anomalies with short phrases, the difference in patterns becomes apparent. In particular, the increase between a phrase of 11 characters and one of 13 shows that longer strings are not much harder for a user to type, but decrease the success rate of an adversary.

>>>>

\author{Noam Zilberstein and Jonathan Chen\thanks{Advised by Nadia Henninger} \\
University of Pennsylvania
}
\maketitle

@@

\section{Introduction}
Using passwords for authentication is an ancient concept that continues to be the commonly accepted standard. For the sake of convenience, many people choose low-strength passwords and reuse them for multiple websites~\cite{Florencio:2007:LSW:1242572.1242661}. This is extremely insecure and begs the question: is there a better way to authenticate users?

Several fixes to this problem are already in place and work by augmenting the password with a secondary authentication factor. Websites that support two factor authentication include Amazon, Facebook, Google, Dropbox, and many others~\cite{2fac}. In each case, the first factor is a password and the second factor is some kind of trusted hardware such as a personal computer or smartphone. When users try to log in from an untrusted machine, they must confirm their identity by typing a code that is sent to a trusted source such as an SMS, email, or phone call. While this process greatly improves the security of a poorly chosen password, it is extremely cumbersome. It also fails if the user does not have access to a trusted piece of hardware. The solution presented in this paper uses trusted hardware that users have at all times: their hands.

@@

A problem arises with this protocol in the case that the database containing users' keystroke data is compromised. In a password model, if the password is leaked, then the user can simply create a new one. However, in an authentication scheme based on keystroke dynamics, if an adversary gains knowledge of keystroke data, then the affected users can never confidently use keystroke dynamics to authenticate themselves again. In order to fix this, as little data as possible should be stored in the database. In fact, the protocol we present is zero-knowledge. This means that users will never at any point in the enrollment or authentication process jeopardize the secrecy of their keystroke timings.

@@

\subsection{Keystroke Dynamics}
\label{sec:bg-keystroke}
Long before computers were invented, written signatures were used to verify people's identities. Signatures are commonly used to confirm the acceptance of contracts and to verify credit card and bank transactions. Typing one's name into a computer is the digital age's moral equivalent to writing it down on a piece of paper. It is not considered a breach of security to see someone's signature written on a piece of paper; the security assumption is that another person cannot copy it to a sufficient level of precision. Similarly, people type their names with high enough frequency that they develop a muscle memory for this process~\cite{monrose:authentication}.

@@

\begin{quote}\small
Peggy would like to convince Victor that two cards are different colors, however Victor is color blind. First, Victor holds up the two cards and asks Peggy to identify the color of each one. Victor then labels the back of the two cards with Peggy's responses, shuffles them behind his back and asks Peggy to identify the color of each card again. If the cards indeed have different colors, then Peggy can easily identify them. However, if the cards are the same color then she can only give an answer that is consistent with the labels with probability $\frac12$. By repeating this process several times, Victor can be sure that Peggy is not lying. However, Victor cannot recreate the proof himself because he is color blind.
\end{quote}

\noindent Relating this to an authentication protocol, the user can be viewed as a prover who knows some secret and the server is a verifier that must confirm that the user knows the secret, but has no access to the secret itself. In the case of keystroke dynamics, the secret is the user's keystroke timings. In a sense users do not ``know'' their timings, but as stated in Section~\ref{sec:bg-keystroke}, they can consistently produce them. If users know the secret then they should always be able to prove their identities, but with high probability an imposter should not be able to prove knowledge of the secret. These conditions relate directly to the example above. If Peggy were cheating, then she would only have a small probability of guessing the labels correctly many times. In an authentication scheme, this failure probability should be negligibly small. In other words, the probability that anyone can prove that they are a certain user without knowing that user's secret is negligible.

@@

However, if the prover does not know the secret, then it would have to produce an alternate value for $c$ such that $g^c\equiv g^{r+bx}\mod p$. The probability of being able to do this is negligible. The protocol is zero-knowledge because, by the discrete log assumption, an adversary can gain no information about $r$ from the value $g^r$, and without $r$ it is impossible to determine $x$ from the value $r + xb$.

@@

\subsection{Fuzzy Extractors}
\label{sec:bg-fuzzy}
Fuzzy extractors are used to extract a uniformly random string $S$ from a noisy input $\mathfrak{B}$, also known as a biometric template. A similar input $\tilde{\mathfrak{B}}$ will produce the exact same string $S$~\cite{conf/csreaSAM/AlvarezEA09}. In the case of keystroke based authentication, $S$ is a user's unique signature which can be extracted from a noisy input. It is then possible to prove knowledge of the secret value $S$ using a zero-knowledge proof. The following construction for a fuzzy extractor is due to \'Alvarez, Encinas, Hern\'andez, and Sanchez-Avila.

@@

\begin{figure}[h!]
\centering
\begin{tikzpicture}[node distance = .25cm and 1cm]
    % Place nodes
    \node [block] (msg) {Message};
    \node [block, right= 3cm of msg] (enc) {Encoded Message};
    \node [decision, right= of enc] (plus) {$-$};
    \node [below= of enc] (hash) {$H$};
    \node [below= of plus] (delta) {$\Delta$};
    \node [above= of plus] (temp) {$\mathfrak{B}$};
    \node [block, above left= of msg] (poly) {$p(x)$};
    \node [left= of poly] (secret) {Secret ($S$)};
    \node [below left= of msg] (rand1) {};
    \node [left= of rand1] (rand) {Random Inputs};
    % Draw edges
    \path [line] (secret) -- (poly);
    \path [line] (rand) -| (poly);
    \path [line] (poly) -| node [midway, above] {$y_1y_2\ldots y_n$} (msg);
    \path [line] (rand) -| node [midway, below] {$x_1x_2\ldots x_n$} (msg);
    \path [line] (msg) -- node [midway, above] {\tiny Reed-Solomon Code} (enc);
    \path [line] (enc) -- (hash);
    \path [line] (enc) -- (plus);
    \path [line] (plus) -- (delta);
    \path [line] (temp) -- (plus);
\end{tikzpicture}
\caption{Enrollment phase for a Fuzzy Extractor}
\label{fig:enroll}
\end{figure}

@@

\begin{figure}[h!]
\centering
\begin{tikzpicture}[node distance = .25cm and .75cm]
    % Place nodes
    \node (temp) {$\tilde{\mathfrak B}$};
    \node [decision, right= of temp] (plus) {$+$};
    \node [above= of plus] (delt) {$\Delta$};
    \node [below = of plus] (reject) {Reject};
    \node [decision, right = of plus] (decide) {$\stackrel?=$};
    \node [above = of decide] (h) {$H$};
    \node [block, right= 2cm of decide] (dec) {Decode $(x_i,y_i)$};
    \node [block, right= 1.5cm of dec] (coeff) {$p(x)$};
    \node [right= of coeff] (s) {$S$};
    % Draw edges
    \path [line] (temp) -- (plus);
    \path [line] (delt) -- (plus);
    \path [line] (plus) -- node [midway, above] {$\mathfrak h$} (decide);
    \path [line] (h) -- (decide);
    \path [line] (decide) -- node [midway, below] {\tiny Reed-Solomon}
      node [near start, above] {\tiny yes} (dec);
    \path [line] (decide) |- node [near start, right] {\tiny no} (reject);
    \path [line] (dec) -- node [midway, above] {\tiny Interpolate} (coeff);
    \path [line] (coeff) -- (s);
\end{tikzpicture}
\caption{Extraction phase for a Fuzzy Extractor}
\label{fig:extract}
\end{figure}

\'Alvarez, Encinas, Hern\'andez, and Sanchez-Avila showed that the above method of fuzzy extraction has a genuine acceptance rate of 89\% and a false acceptance rate of around 1\% on iris scan authentication~\cite{conf/csreaSAM/AlvarezEA09}. While a different biometric such as keystroke dynamics would undoubtably produce different error rates, prior work shows that this fuzzy extractor has potential to accurately distinguish between people. The genuine acceptant rate of 89\% is fairly good; it implies that the expected number of trials people will need to perform in order to authenticate themselves is $\frac{100}{89}\approx 1.12$. In the context of the keystroke dynamics protocol, this is convenient enough to satisfy even a lazy user. In most instances, the user will only have to type his name once, but occasionally a second trial may be required.

@@

We wish to examine the feasibility of keystroke based authentication from both a theoretical and empirical perspective. Two protocols are presented. The first protocol is extremely insecure, but acts as a proof of concept and comparison baseline in terms of accuracy and security. This protocol has been implemented as a simple web app. The second protocol is a zero-knowledge version of the first one that relies on an extractor to obtain a secret from a noisy input. The secret is never stored or communicated across a network; instead knowledge of the secret is proven using a zero-knowledge protocol. This protocol has been implemented using a rounding extractor and the feasibility of using the fuzzy extractor that is described in Section~\ref{sec:bg-fuzzy} is also examined.

When discussing these protocols, the words \textit{prover} and \textit{client} are used almost interchangeably as are the words \textit{verifier} and \textit{server}. The client is an implementation of a prover and the server is an implementation of a verifier. The client is naturally more `powerful' than the server as the client knows the secret keystroke timings (or can at least extract them from a noisy input). Both the regular and the zero-knowledge versions of the web app have two capabilities: account creation and login. The account creation phase requires users to type their names in a text box. The keystroke timings are recorded and communicated to the server. The server then stores some data about the keystrokes in a database. When users try to log in, they also simply types their names in a text box. The names that are typed are looked up in the database and then their keystroke timings are compared against the ones that are stored. Note that the word \emph{compare} is used loosely here; in the zero-knowledge protocol, no direct comparison occurs per se. Instead, the difference between the two signatures is measured by some specialized technique. The user experience is identical (aside from login accuracy) in the regular and zero-knowledge versions, although the implementation is quite different.

@@

\subsection{The Zero-Knowledge Protocol}
\label{sec:zkp-prot}
The zero-knowledge protocol builds on top of the basic protocol with the added constraint that no meaningful information about the keystroke vector is ever transferred across the network. This also means that the representation of the keystroke vector that is stored in the database does not leak any information about the keystroke vector itself. If the database were completely compromised, there would thus be no breach in security; the information in the database can be used to verify a user's identity and nothing more.

@@

The authentication protocol is shown in Figure~\ref{fig:zk_auth}. Since the client in this implementation is a web page that may be accessed from any computer, information about a particular user cannot be stored and thus the value of $g$ must be provided by the server after discovering the user's name. Aside from this caveat, the protocol is identical to the zero-knowledge proof presented in Section~\ref{sec:bg-zkp}.

@@

Table \ref{table:res-basic} contains the results from this experiment en the Basic Protocol implementation that is described in Section~\ref{sec:basic-prot}. Trials were conducted between 20 and 40 times with varying word lengths. The values of the timings were compared directly, allowing a total error of $\epsilon=50$. The value $\epsilon'$ in Table~\ref{table:res-basic} indicates the average error value that the target user scored. Some trends arise from these results. As word length increases, typing signatures become more unique and therefore adversary forgery success rate drops significantly. However, this also affects target users' ability to recreate their typing signature consistently. There appears to be a tradeoff between security and convenience which is due to choosing a constant $\epsilon$. Longer keystroke vectors amortize the allowed error over a larger number of keystrokes, therefore each keystroke must be more exact.

@@

Table~\ref{table:res-zkp} documents repeated test trials with $\delta=0.05$. Trials were conducted between 20 and 40 times with varying word lengths. The trend of increasing successful authentications with increasing word length is apparent. This is most likely attributed to the fact that typing patterns are more distinct with longer pieces of text. Conversely, imposters are also more likely to successfully impersonate the target user with a longer text, since the scheme does not measure the accumulated difference in time. In other words, an imposter's speed could be very far from the expected timings on many of the strokes, but close on others. In this case, the imposter may still be authenticated because the protocol only cares \textit{how many} matches occurred, not \textit{how much} the totals differ. With shorter words, the implementation is very unforgiving.  Since there are fewer keystrokes to average out the mistakes, timing is extremely sensitive.  In practice, this is less of a concern since virtually no one has a combined first name and last name of less than three characters. There exists a non-trivial difference between the success rates of target users and their impersonators. This affirms the fact that keystroke dynamics are variable enough and have sufficient entropy to be used as authentication tools. Barring anomalies with short phrases, the difference in patterns becomes apparent. In particular, the increase between a phrase of 11 characters and one of 13 shows that longer strings are not much harder for a user to type, but decrease the success rate of an adversary.
